---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(callme)
```


* Scope
  * Calling C (not C++)
    * Examples aiming for clarity/consistency. Faster methods/shortcuts may be available.
  * using the .Call interface (contrast with .C, .Fortran)
  * Other methods of calling inline: `cpp11::cpp_source()`
  * References
    * Advanced R book (hadley)
    * Writing R Extensions
    * R source code
* General Approach
  * Be comfortable programming in C
  * Each value has to be unpacked from a generic SEXP type to an R type
  * Need to create R values and PROTECT them from being garbage collected
  * Release them prior to return
  * types: int, real, logical, character, list, data.frame
    * Unpack and use.  Create a new one in C and return.
    * use type2str(TYPEOF()) to show type name
    * switch() statement on TYPEOF()
    * complex example which takes a few arguments and returns a data.frame
  * Special print methods:  Rprintf, message, warning, error
* Accessing a third-party library  (use zlib as an example maybe)
* Accessing RNG
* Calling R from C

# Introduction

`{callme}` is a package for easily compiling inline C code for use within R.

You should definitely read Hadley's [Advanced R Book](http://adv-r.had.co.nz/C-interface.html)
as it has a great chapter or R's interface to C.

## Calling compiled code in R

There are a number of possible ways to call compiled code in R:

* `.Call()` - as used by this package
* `.External()` - an alternative to `.Call()` with a different way of passing
  arguments. 
. `.C()` a simple C interface.  Code compiled with this package is not 
  compatible with this interface.
. `.Fortran()` an interface to Fortran code.  Definitely not supported or used within this package.

This package only supports calling via the `.Call()` interface.

## Writing code compatible with `.Call()`

The following code adds two vectors of floating point values and returns
the result i.e. `a * b`.

```
SEXP add(SEXP a, SEXP b) {
  
  // Sanity checks
  if (length(a) != length(b)) {
    error("'a' and 'b' must be the same length");
  }
  
  // Get a pointer to the actual numeric data in 'a' and 'b'
  double *ap = REAL(a);
  double *bp = REAL(b);
  
  // Allocate a new R object 'res'
  int N = length(a);
  SEXP res = PROTECT(allocVector(REALSXP, N));
  
  // Get a pointer to the actual numeric data in 'res'
  double *resp = REAL(res);
  
  // Perform some standard C-style operations
  for (int i = 0; i < N; i++) {
    resp[i] = ap[i] + bp[i];
  }
  
  // Unwind and data protection and return the R result
  UNPROTECT(1);
  return res;
}
```

```{r echo=FALSE, eval=TRUE}
code <- r"(
SEXP add(SEXP a, SEXP b) {
  
  // Sanity checks
  if (length(a) != length(b)) {
    error("'a' and 'b' must be the same length");
  }
  
  // Get a pointer to the actual numeric data in 'a' and 'b'
  double *ap = REAL(a);
  double *bp = REAL(b);
  
  // Allocate a new R object 'res'
  int N = length(a);
  SEXP res = PROTECT(allocVector(REALSXP, N));
  
  // Get a pointer to the actual numeric data in 'res'
  double *resp = REAL(res);
  
  // Perform some standard C-style operations
  for (int i = 0; i < N; i++) {
    resp[i] = ap[i] + bp[i];
  }
  
  // Unwind and data protection and return the R result
  UNPROTECT(1);
  return res;
}
)"
callme::compile(code)
```


The following sections outline the sequence of steps in writing a C
function to call from R, and is the template followed for the vignettes in this
package.

* Function signature
* Sanity checking
* Unpack the R objects into their C equivalents
* Allocate a new R object to use as the result
* Standard C code
* Tidy and return


#### Function signature

All functions callable from R must return a `SEXP`, and only take arguments 
which are of type `SEXP` (or maybe just `void` if there are no arguments).

The `SEXP` type is an `S-Expression`.  Every value in R is an `SEXP`, and 
there is information stored within the type to indicate what data is stored 
within it.

The `TYPEOF()` macro and `type2char()` function within C will be useful to 
identify what sort of data is in the SEXP.

```{r}
callme::compile(r"(
SEXP what_sexp_is_this(SEXP x) {
  Rprintf("SEXPTYPE: %i = %s\n", TYPEOF(x), type2char(TYPEOF(x)));
  return R_NilValue;
}                
)", invisible = TRUE)

what_sexp_is_this(1L)
what_sexp_is_this(TRUE)
what_sexp_is_this(c(1.1, 2.2))
what_sexp_is_this(list(1, 2, 3))
what_sexp_is_this("hello")
what_sexp_is_this(mtcars)
what_sexp_is_this(mean)
```

The full list of `SEXP` types are in [Rinternals.h](https://github.com/wch/r-source/blob/trunk/src/include/Rinternals.h)

```
#define NILSXP	     0	  /* nil = NULL */
#define SYMSXP	     1	  /* symbols */
#define LISTSXP	     2	  /* lists of dotted pairs */
#define CLOSXP	     3	  /* closures */
#define ENVSXP	     4	  /* environments */
#define PROMSXP	     5	  /* promises: [un]evaluated closure arguments */
#define LANGSXP	     6	  /* language constructs (special lists) */
#define SPECIALSXP   7	  /* special forms */
#define BUILTINSXP   8	  /* builtin non-special forms */
#define CHARSXP	     9	  /* "scalar" string type (internal only)*/
#define LGLSXP	    10	  /* logical vectors */
/* 11 and 12 were factors and ordered factors in the 1990s */
#define INTSXP	    13	  /* integer vectors */
#define REALSXP	    14	  /* real variables */
#define CPLXSXP	    15	  /* complex variables */
#define STRSXP	    16	  /* string vectors */
#define DOTSXP	    17	  /* dot-dot-dot object */
#define ANYSXP	    18	  /* make "any" args work.
			     Used in specifying types for symbol
			     registration to mean anything is okay  */
#define VECSXP	    19	  /* generic vectors */
#define EXPRSXP	    20	  /* expressions vectors */
#define BCODESXP    21    /* byte code */
#define EXTPTRSXP   22    /* external pointer */
#define WEAKREFSXP  23    /* weak reference */
#define RAWSXP      24    /* raw bytes */
#define OBJSXP      25    /* object, non-vector  */
#define S4SXP       25    /* same as OBJSXP, retained for back compatability */

/* used for detecting PROTECT issues in memory.c */
#define NEWSXP      30    /* fresh node created in new page */
#define FREESXP     31    /* node released by GC */

#define FUNSXP      99    /* Closure or Builtin or Special */
```



#### Sanity checking

The `SEXP` type can contain any R object, so when passed in to a C function from 
R, it is necessary to confirm that it is sane i.e.

* Is it the right type of data?
* Is it the right length?
* Do all objects together make sense? e.g. must they all be the same length?

Some checks will need to be explicit.  In the example, the `a` and `b` arguments
are checked if they are the same length and an error is raised if not.  (`error()` is
the C equivalenth of R's `stop()`)

```
// Sanity checks
if (length(a) != length(b)) {
  error("'a' and 'b' must be the same length");
}
```

If this check was skipped, then later in the example the expression `a[i] * b[i]`
is possibly going to try and access memory that is out-of-bounds!


There are some checks which are implicit in that R will complain if you try 
to use a value in an inappropriate way.  For example, if a character vector
is passed in as an argument to `add()`, the `REAL()` macro will try to 
extract numeric data from the string and fail:

```{r error = TRUE}
add(1.0, 2.0)     # Good: both types are floating point numeric values
add(1.0, "hello") # Bad : Can't treat "hello" as a number 
```



#### Unpack the R objects into their C equivalents

* `LOGICAL()` extracts the `int32_t` vector with values `0` and `1` from the SEXP
* `INTEGER()` extracts the `int32_t` vector from the SEXP
* `REAL()` extracts the `double` pointer
* `RAW()` extracts the `uint8_t` pointer
* `COMPLEX()` extracts the `double` pointer to the interleaved real/imaginary components
* `VECTOR_ELT()` extracts an element from a R `list`, or column from an R `data.frame`
* `STRING_ELT()` extracts an element from an R character vector


```{r}
compile(r"(
#include <stdint.h>
SEXP unpack_integers_from_list(SEXP ll) {
  
  // sanity check
  if (TYPEOF(ll) != VECSXP) {
    error("'ll' must a list");
  }

  // iterate over the list
  for (int i = 0; i < length(ll); i++) {
    SEXP elem = VECTOR_ELT(ll, i);
    if (TYPEOF(elem) == INTSXP) {
       Rprintf("Integer    : ");
       int32_t *intp = INTEGER(elem);
       for (int j = 0; j < length(elem); j++) {
          Rprintf("%i ", intp[j]);
       }
       Rprintf("\n");
    } else {
      Rprintf("Non-integer: %s\n", type2char(TYPEOF(elem)));
    }
  }

  return R_NilValue;
})", invisible = TRUE)

unpack_integers_from_list(list(1:3, "hello", 1.1, TRUE))
```







#### Allocate a new R object to use as the result

Standard R types may be created with a call to `allocVector()`.

Because we are operating within C, and the object is not assigned to an R 
variable, it is necessary to tell R that we are using the object.  This 
is done with a `PROTECT()` call, and stops R from trying to free the memory 
while we are using it by calling the garbage collector.

In general, every R object created within C should be wrapped in a `PROTECT()`.

When we leave the function, tally the calls to `PROTECT()` and undo them with
`UNPROTECT(n)`.  



```{r}
callme::compile(r"(
SEXP alloc_integer(SEXP n) {
  int len = INTEGER(n)[0];
  SEXP res = PROTECT(allocVector(INTSXP, len));
  int *resp = INTEGER(res);
  for (int i = 0; i < len; i++) {
     resp[i] = i;
  }
  UNPROTECT(1);
  return res;
})")

alloc_integer(10L)
```



#### Standard C code

In the middle of all the `SEXP`, sanity checking, unpacking and PROTECTing,
there's often an identifiable bit of just plain C code.  This is usually
the entire purpose of calling C from R

In the above example this core C code is just looping over two C arrays
of doubles and summing the result.

```
for (int i = 0; i < N; i++) {
  resp[i] = ap[i] + bp[i];
}
```


#### Tidy and return

* `UNPROTECT()` any objects you `PROTECT()`ed
* Return an `SEXP` object.
* Any other `SEXP` objects allocated within the function but not returned will
  be garbage collected later by R.



