---
title: "SEXP objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SEXP objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::knit_engines$set(callme = callme::callme_engine)
library(callme)
```

```{css, echo=FALSE}
.callme         { background-color: #E3F2FD; }
pre.callme span { background-color: #E3F2FD; }
```

# SEXP objects

All functions callable from R must return a `SEXP`, and only take arguments 
which are of type `SEXP` (or just `void` if there are no arguments).

The `SEXP` type is an `S-Expression`.  Every value in R is an `SEXP`, and 
there is information stored within the type to indicate what data is stored 
within it.

The `TYPEOF()` macro and `type2char()` function within C will be useful to 
identify what sort of data is in the SEXP.


## Listing of all SEXP types

The full list of `SEXP` types are in [Rinternals.h](https://github.com/wch/r-source/blob/trunk/src/include/Rinternals.h)

```
#define NILSXP	     0	  /* nil = NULL */
#define SYMSXP	     1	  /* symbols */
#define LISTSXP	     2	  /* lists of dotted pairs */
#define CLOSXP	     3	  /* closures */
#define ENVSXP	     4	  /* environments */
#define PROMSXP	     5	  /* promises: [un]evaluated closure arguments */
#define LANGSXP	     6	  /* language constructs (special lists) */
#define SPECIALSXP   7	  /* special forms */
#define BUILTINSXP   8	  /* builtin non-special forms */
#define CHARSXP	     9	  /* "scalar" string type (internal only)*/
#define LGLSXP	    10	  /* logical vectors */
/* 11 and 12 were factors and ordered factors in the 1990s */
#define INTSXP	    13	  /* integer vectors */
#define REALSXP	    14	  /* real variables */
#define CPLXSXP	    15	  /* complex variables */
#define STRSXP	    16	  /* string vectors */
#define DOTSXP	    17	  /* dot-dot-dot object */
#define ANYSXP	    18	  /* make "any" args work.
			     Used in specifying types for symbol
			     registration to mean anything is okay  */
#define VECSXP	    19	  /* generic vectors */
#define EXPRSXP	    20	  /* expressions vectors */
#define BCODESXP    21    /* byte code */
#define EXTPTRSXP   22    /* external pointer */
#define WEAKREFSXP  23    /* weak reference */
#define RAWSXP      24    /* raw bytes */
#define OBJSXP      25    /* object, non-vector  */
#define S4SXP       25    /* same as OBJSXP, retained for back compatability */

/* used for detecting PROTECT issues in memory.c */
#define NEWSXP      30    /* fresh node created in new page */
#define FREESXP     31    /* node released by GC */

#define FUNSXP      99    /* Closure or Builtin or Special */
```

## Code example: Print the SEXP type of an object

```{callme}
#| invisible = TRUE
SEXP what_sexp_is_this(SEXP x) {
  Rprintf("SEXPTYPE: %i = %s\n", TYPEOF(x), type2char(TYPEOF(x)));
  return R_NilValue;
}                
```


```{r}
what_sexp_is_this(1L)
what_sexp_is_this(TRUE)
what_sexp_is_this(c(1.1, 2.2))
what_sexp_is_this(list(1, 2, 3))
what_sexp_is_this("hello")
what_sexp_is_this(mtcars)
what_sexp_is_this(mean)
```


## Using `TYPEOF()` and `is*()` functions to determine type

There are many specialised functions for detecting types and data structures.

A non-exhaustive example of type testing is shown here:

```{callme}
#| invisible=TRUE
SEXP show_type(SEXP x) {
  Rprintf("----- Using TYPEOF() -----------\n");
  if (TYPEOF(x) == INTSXP ) Rprintf("INTSXP\n");
  if (TYPEOF(x) == REALSXP) Rprintf("REALSXP\n");
  if (TYPEOF(x) == STRSXP ) Rprintf("STRSXP\n");
  if (TYPEOF(x) == RAWSXP ) Rprintf("RAWSXP\n");
  if (TYPEOF(x) == VECSXP ) Rprintf("VECSXP\n");
  if (TYPEOF(x) == LGLSXP ) Rprintf("LGLSXP\n");
  Rprintf("----- Using is*() -----------\n");
  if (isInteger(x)) Rprintf("INTSXP\n");
  if (isReal(x))    Rprintf("REALSXP\n");
  if (isString(x))  Rprintf("STRSXP\n");
  if (isNewList(x)) Rprintf("List\n");
  if (isFactor(x))  Rprintf("Factor\n");
  if (isArray(x))   Rprintf("Array\n");
  if (isMatrix(x))  Rprintf("Matrix\n");

  return R_NilValue;
}
```

```{r}
show_type(1)
show_type(iris$Species)
show_type(matrix(1, 2, 2))
show_type(matrix('a', 2, 2))
```







# Unpacking SEXP objects to C data



* `LOGICAL()` extracts the `int32_t` vector with values `0` and `1` from the SEXP
* `INTEGER()` extracts the `int32_t` vector from the SEXP
* `REAL()` extracts the `double` pointer
* `RAW()` extracts the `uint8_t` pointer
* `COMPLEX()` extracts the `double` pointer to the interleaved real/imaginary components
* `VECTOR_ELT()` extracts an element from a R `list`, or column from an R `data.frame`
* `STRING_ELT()` extracts an element from an R character vector

## Integers

```{callme}
#| invisible = TRUE
SEXP unpack_integer(SEXP vec_) {

  if (!isInteger(vec_)) error("Expecting integer, but got a '%s'", type2char(TYPEOF(vec_)));

  int *vec = INTEGER(vec_);

  Rprintf("First integer (Method 1): %i\n", vec[0]);
  Rprintf("First integer (Method 2): %i\n", asInteger(vec_));

  return R_NilValue;
}
```

```{r}
unpack_integer(1:3)
```


## Real

```{callme}
#| invisible = TRUE
SEXP unpack_real(SEXP vec_) {

  if (!isReal(vec_)) error("Expecting real, but got a '%s'", type2char(TYPEOF(vec_)));

  double *vec = REAL(vec_);

  Rprintf("First number (Method 1): %f\n", vec[0]);
  Rprintf("First number (Method 2): %f\n", asReal(vec_));

  return R_NilValue;
}
```

```{r}
unpack_real(c(1.23, 4.56))
```

## Logical

```{callme}
#| invisible = TRUE
SEXP unpack_logical(SEXP vec_) {

  if (!isLogical(vec_)) error("Expecting logical, but got a '%s'", type2char(TYPEOF(vec_)));

  double *vec = LOGICAL(vec_);

  Rprintf("First logical (Method 1): %i\n", vec[0]);
  Rprintf("First logical (Method 2): %i\n", asLogical(vec_));

  return R_NilValue;
}
```

```{r}
unpack_logical(c(T, F))
```


## Character Vector

```{callme}
#| invisible = TRUE
SEXP unpack_character(SEXP vec_) {

  if (!isString(vec_)) error("Expecting character vector, but got a '%s'", type2char(TYPEOF(vec_)));

  Rprintf("First string: %s\n", CHAR(STRING_ELT(vec_, 0)));

  return R_NilValue;
}
```

```{r}
unpack_character(c('hello', 'there'))
```


## List

```{callme}
#| invisible = TRUE
SEXP unpack_list(SEXP vec_) {
  
  if (!isNewList(vec_)) error("Not a 'list'");

  SEXP elem = VECTOR_ELT(vec_, 0);
  Rprintf("First list element is of type: '%s'\n", type2char(TYPEOF(elem)));

  return R_NilValue;
}
```

```{r}
unpack_list(list(1.23, "hello"))
```


# Unpacking scalar values

There are specialist functions for unpacking scalar values.  These
behave similar to their R equivalents to do type conversion, but
results can be a little more unpredictable


Note below, that an `NA_integer_` value is equivalent to the smallest signed
32-bit integer in C (i.e. `-(2^31) = -2147483648`)

```{callme}
#| invisible = TRUE
SEXP unpack_integer_scalar(SEXP vec_) {

  Rprintf("First integer: %i\n", asInteger(vec_));
  Rprintf("First real   : %i\n", asReal(vec_));
  Rprintf("First logical: %i\n", asLogical(vec_));

  return R_NilValue;
}
```

```{r}
unpack_integer_scalar(1L)
unpack_integer_scalar(1.23)
unpack_integer_scalar("1")
unpack_integer_scalar("hello")
```




# Advanced example 

Examine all values in the provided list, and if the value is an integer vector
then print all the numbers, otherwise just print the type of element:


```{callme}
#| invisible = TRUE
#include <stdint.h>
SEXP unpack_integers_from_list(SEXP ll) {
  
  // sanity check
  if (TYPEOF(ll) != VECSXP) {
    error("'ll' must a list");
  }

  // iterate over the list
  for (int i = 0; i < length(ll); i++) {
    SEXP elem = VECTOR_ELT(ll, i);
    if (TYPEOF(elem) == INTSXP) {
       Rprintf("Integer    : ");
       int32_t *intp = INTEGER(elem);
       for (int j = 0; j < length(elem); j++) {
          Rprintf("%i ", intp[j]);
       }
       Rprintf("\n");
    } else {
      Rprintf("Non-integer: %s\n", type2char(TYPEOF(elem)));
    }
  }

  return R_NilValue;
}
```

```{r}
unpack_integers_from_list(list(1:3, "hello", 1.1, TRUE))
unpack_integers_from_list(iris)
```















