---
title: "SEXP objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SEXP objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::knit_engines$set(callme = callme:::callme_engine)
library(callme)
```

```{css, echo=FALSE}
.callme         { background-color: #E3F2FD; }
pre.callme span { background-color: #E3F2FD; }
```

## SEXP objects

All functions callable from R must return a `SEXP`, and only take arguments 
which are of type `SEXP` (or just `void` if there are no arguments).

The `SEXP` type is an `S-Expression`.  Every value in R is an `SEXP`, and 
there is information stored within the type to indicate what data is stored 
within it.

The `TYPEOF()` macro and `type2char()` function within C will be useful to 
identify what sort of data is in the SEXP.


## Listing of all SEXP types

The full list of `SEXP` types are in [Rinternals.h](https://github.com/wch/r-source/blob/trunk/src/include/Rinternals.h)

```
#define NILSXP	     0	  /* nil = NULL */
#define SYMSXP	     1	  /* symbols */
#define LISTSXP	     2	  /* lists of dotted pairs */
#define CLOSXP	     3	  /* closures */
#define ENVSXP	     4	  /* environments */
#define PROMSXP	     5	  /* promises: [un]evaluated closure arguments */
#define LANGSXP	     6	  /* language constructs (special lists) */
#define SPECIALSXP   7	  /* special forms */
#define BUILTINSXP   8	  /* builtin non-special forms */
#define CHARSXP	     9	  /* "scalar" string type (internal only)*/
#define LGLSXP	    10	  /* logical vectors */
/* 11 and 12 were factors and ordered factors in the 1990s */
#define INTSXP	    13	  /* integer vectors */
#define REALSXP	    14	  /* real variables */
#define CPLXSXP	    15	  /* complex variables */
#define STRSXP	    16	  /* string vectors */
#define DOTSXP	    17	  /* dot-dot-dot object */
#define ANYSXP	    18	  /* make "any" args work.
			     Used in specifying types for symbol
			     registration to mean anything is okay  */
#define VECSXP	    19	  /* generic vectors */
#define EXPRSXP	    20	  /* expressions vectors */
#define BCODESXP    21    /* byte code */
#define EXTPTRSXP   22    /* external pointer */
#define WEAKREFSXP  23    /* weak reference */
#define RAWSXP      24    /* raw bytes */
#define OBJSXP      25    /* object, non-vector  */
#define S4SXP       25    /* same as OBJSXP, retained for back compatability */

/* used for detecting PROTECT issues in memory.c */
#define NEWSXP      30    /* fresh node created in new page */
#define FREESXP     31    /* node released by GC */

#define FUNSXP      99    /* Closure or Builtin or Special */
```

## Code example: Print the SEXP type of an object

```{callme}
#| invisible = TRUE
SEXP what_sexp_is_this(SEXP x) {
  Rprintf("SEXPTYPE: %i = %s\n", TYPEOF(x), type2char(TYPEOF(x)));
  return R_NilValue;
}                
```


```{r}
what_sexp_is_this(1L)
what_sexp_is_this(TRUE)
what_sexp_is_this(c(1.1, 2.2))
what_sexp_is_this(list(1, 2, 3))
what_sexp_is_this("hello")
what_sexp_is_this(mtcars)
what_sexp_is_this(mean)
```






